\section{Results}
\label{sec:results}

First of all, we used our toolkit to attack our straightforward C++ implementation of AES, here referred to as {\tt naiveAES}, which was supposed to serve just as a proof-of-concept; see results in Section~\ref{sec:naiveaes}. Note that Section~\ref{sec:naiveaes} does not fully cover many details, these are given later in subsequent sections.

Then we performed the most interesting attack that Bos et al.\ \cite{bos2015differential} did -- we successfully attacked Klinec's C++ implementation \cite{klinec2013implementation} of WBAES by Chow et al.\ \cite{chow2002aes}, here referred to as {\tt KlinecWBAES}; see results and comments in Section~\ref{sec:klinecwbaes}.

%_% bačinská
Finally in Section~\ref{sec:bacinskawbaes}, we studied possible usage of the attack against a white-box implementation of a cipher derived from WBAES, introduced by Bačinská \cite{bacinska2015white}, here referred to as {\tt BacinskaWBAES+}.

\begin{note}
\label{note:expdes}
	In our results, we only attacked implementations with known source codes and known key. The case of unknown key will be discussed further in Section~\ref{sec:blindattack}.
\end{note}


% ==============================================================================
% ===   n a i v e A E S                                                      ===
% ==============================================================================

\subsection{\tt naiveAES}
\label{sec:naiveaes}

In order to prove the concept of both SCA algorithms presented in this thesis, we first attacked our straightforward AES implementation {\tt naiveAES} with both of them. Both attacks are expected to work, since {\tt naiveAES} stores all intermediate results that are required for each attack. For this reason, we captured contents of all $1$-byte memory writes into our memory traces, and used output of the first SBox as our target. In both cases, we treated strong candidates differently (see Note~\ref{note:strong} for what a strong candidate is).

\subsubsection{CPA Attack against {\tt naiveAES}}
	
	First we attacked {\tt naiveAES} with the CPA attack (i.e., Algorithm~\ref{alg:cpa}) using $64$, $96$ and $128$ traces, respectively; see results in Table~\ref{tab:naiveaescpa} and Note~\ref{note:tabvals} for its description.
	
	\newpage   %_%
	
	\begin{table}[h]
		\begin{center}
		\input{./tables/3_2_1-naiveAES-CPA.tex}
		\end{center}
	\caption{CPA attack against {\tt naiveAES} using different number of traces. Percentual gap of the best candidate and rank of the correct candidate is given, for each key byte and each number of traces. The rank ranges from $0$, while $0$ (i.e., the top position) is replaced with $\blacksquare$ or {\weak$\blacksquare$} for strong or weak candidate, respectively, in order to emphasize successful attack.}
	\label{tab:naiveaescpa}
	\end{table}
	
	\begin{note}
	\label{note:tabvals}
		In our tables, two values will be given: percentual gap of the best candidate to the second best one together with rank of the correct candidate, for each key byte and each number of traces. Note that the rank ranges from $0$ to $255$, while $0$ is replaced with $\blacksquare$ or {\weak$\blacksquare$} for strong or weak candidate, respectively, in order to emphasize successful attack. See the following example for a hint.
	\end{note}
	
	\begin{example}
	\label{ex:gaprank}
		If we looked at the \nth{5} bit back in Table~\ref{tab:gap} in Example~\ref{ex:gap}, these two values would be $\frac{0.3970-0.3884}{0.3970}\cdot100\%\approx2.2\%$ and $45$, respectively.
	\end{example}
	
	It appears that we need hundreds of traces in order to attack even an unprotected implementation successfully. Note that in the CPA attack, we only have a single attack target (for each key byte), which is the Hamming weight of the first SBox output (cf.\ $8$ target bits in the bitwise DPA). However, this attack served just as a proof of the concept.
	
	\begin{note}
	\label{note:target}
		The term {\em target of the attack} or simply {\em target} will be used heavily in the following text, let us describe it a bit more.
		
		Target is a function $T$ that inputs a key byte $K$ together with the respective plaintext byte $P$, and outputs a byte (for now), or a single bit (later), e.g., $T(K,P) = S(K+P)$ on Lines~\ref{line:s0} and \ref{line:s1} of Algorithm~\ref{alg:bitwisedpa}. Target bit is just a specific bit of target's output. According to the value of target bit, traces are split into the sets $S_0$ and $S_1$, respectively.
	\end{note}

\subsubsection{Bitwise DPA Attack against {\tt naiveAES}}
	
	Afterwards, we performed the bitwise DPA attack (i.e., Algorithm~\ref{alg:bitwisedpa}) against {\tt naiveAES} using $24$ and $32$ memory traces, respectively; see results in Table~\ref{tab:naiveaesdca}. The table shows amount of leaking target bits out of $8$ of them.
	
	\begin{note}
	\label{note:unprotect}
		In case of the correct candidate, the maximum value of the difference of means (which is what the bitwise DPA looks for) is {\em always} equal to $1.0$, since there is always a position in the traces where given intermediate product -- here a bit of the first SBox output -- appears. Therefore once there appears a non-zero gap for the first time, we certainly get the correct candidate and do not need to care about its strength. Obviously, the attack can be aborted at such moment.
	\end{note}
	
	\begin{table}[h]
		\begin{center}
		\input{./tables/3_2_2-naiveAES-DCA.tex}
		\end{center}
	\caption{Number of target bits that gave single candidate with difference of means equal to $1.0$ using the bitwise DPA attack against {\tt naiveAES}.}
	\label{tab:naiveaesdca}
	\end{table}
	
	Using $24$ traces has mostly shown to be sufficient, and with $32$ traces, the attack was successful for every target bit. Note that these numbers are only illustrative -- it could possibly happen that many more traces would keep two candidates on the top, both with the difference of means equal to $1.0$.
	
	We presented these results just to give a comparison to the previous CPA attack and to the bitwise DPA attack against a white-box protected implementation.


% ==============================================================================
% ===   K l i n e c W B A E S                                                ===
% ==============================================================================

\subsection{\tt KlinecWBAES}
\label{sec:klinecwbaes}

The most interesting results came with attacking Klinec's implementation \cite{klinec2013implementation} of WBAES by Chow et al.\ \cite{chow2002aes}, here denoted as {\tt KlinecWBAES}. Note that we will only consider this particular implementation in the rest of this thesis, except for Section~\ref{sec:bacinskawbaes}.

First we tried to attack {\tt KlinecWBAES} with the CPA attack, too, but it did not break any key byte at all. Therefore the rest of our results focuses exclusively on the bitwise DPA attack.

\subsubsection{Bitwise DPA Attack against {\tt KlinecWBAES}}
	
	We used $1024$ traces for the attack; see results in Table~\ref{tab:klinecsbox}. In this table, we give the same values as previously in Table~\ref{tab:naiveaescpa}, see Note~\ref{note:tabvals}.
	
	\afterpage{
		%~ \clearpage   % To flush out all floats, might not be what you want
		\begin{landscape}
		\begin{table}[h]
			\begin{center}
			\input{./tables/3_2_3-klinec-sbox.tex}
			\end{center}
		\caption{Bitwise DPA attack against {\tt KlinecWBAES} using $1024$ traces. Percentual gap of the best candidate and rank of the correct candidate is given, for each key byte and each number of traces; see Note~\ref{note:tabvals} for a full description.}
		\label{tab:klinecsbox}
		\end{table}
		\end{landscape}
	}
	
	On average, there are about $31\%$ of successful strong candidates and about $36\%$ of both weak and strong ones.
	
	Compared to the results of Bos et al., we have so far successfully reproduced one half of their attack. The second half is given in Section~\ref{sec:rijinv}, because it requires some improvements to be introduced.
	
	\begin{note}
	\label{note:tailrank}
		As already noted by Bos et al., we can see remarkably many tail ranks (e.g., between $250$ and $255$) in results in Table~\ref{tab:klinecsbox}.
	\end{note}


% ==============================================================================
% ===   B a c i n s k a W B A E S +                                        ===
% ==============================================================================

%_% bačinská
\subsection{\tt BacinskaWBAES+}
\label{sec:bacinskawbaes}

Bačinská introduced a variant of WBAES in \cite{bacinska2015white}, here denoted as {\tt BacinskaWBAES+}. However, her cipher is not compliant with the standard AES anymore -- SBoxes of {\tt BacinskaWBAES+} are key-dependent, i.e., different from those of the standard AES. Note that the output of the first standard SBox is what the attack is targeting, therefore there is nothing to be successfully attacked. Finally, the attack did not work, as expected.

The attack could be most probably ``customized'' for {\tt BacinskaWBAES+}. However, there is no practical interest in non-standard ciphers, hence we omitted it.