\chapter{Future Work}
\label{chap:future}

\section{Another Interesting AES White-Box Implementation}
\label{sec:magicimpl}

Bos et al.\ \cite{bos2015differential} mention another interesting AES white-box implementation in their paper. This particular implementation by Eloi Vanderbéken was originally published only as a Windows binary executable challenge at NoSuchCon 2013 conference\footnote{\link{http://www.nosuchcon.org/2013/}. Accessed 2016-04-15.}, where nobody was able to recover the key. Note that this implementation uses an externally-encoded variant of AES (as outlined in Section~\ref{sec:constrwbaes}, in particular in Equation~\ref{eq:extenc}), hence it is {\em not compliant with AES anymore}. It follows that any SCA-like attack is impossible, since the internal AES handles encoded inputs, where the encoding is unknown.

Soon after the conference, the author released source codes\footnote{\link{http://pastebin.com/MvXpGZts}. Accessed 2016-04-15.}, which allow one to avoid reverse engineering of the binary. Moreover, it allows one to modify the code to run natively on Linux. Philippe Teuwen, a co-author of \cite{bos2015differential}, studied this implementation deeper, and finally was able to recover the key \cite{teuwen2015nscwriteups}.

In his write-up, Teuwen also describes the structure, which appears to avoid usage of $4$-bit random bijections (as used in WBAES by Chow et al.\ \cite{chow2002aes}) and uses $8$-bit ones instead. Note that even without nonlinearity check, it is far less likely that any output bit of a random $8$-bit bijection would provide a reasonable level of linearity compared to a random $4$-bit bijection, as outlined in Section~\ref{sec:attempt}.

However, we did not either modify the source code to make it AES-compliant, nor did run the attack, hence it remains as a possible topic for further research.


\section{Miscellaneous Remarks}

During writing of this thesis, there emerged a couple of unresolved issues. Here we give a short list of those, which we think would deserve further attention, in the order of their significance we granted them.

%?% všechno?
\begin{description}
	\item[(Semi-)automatic leakage locator.] In Section~\ref{sec:leakpoint}, we mentioned a successful attempt to find the leaking position in source code. A (semi-)automatic tool would be definitely appreciated in white-box cryptography design.
	\item[Adaptive bound.] In Remark~\ref{rem:attimpr}, we outlined a possible enhancement of our attack with adaptive bound, which would certainly accelerate the attack.
	\item[(Non)uniformity in leaking bits.] Leaking bits, as introduced in Note~\ref{note:leakbit}, indicate quite a weird behavior in case of the attack against memory traces; this is captured in histogram in Figure~\ref{fig:leakbitall}. However, this surprisingly did not happen when attacking outputted values, see Note~\ref{note:leakbituniform}. We did not find any explanation for this.
\end{description}
